---
title: "Plotting Growth Curves"
author: "UZH n UU: Who is the author here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=TRUE, echo=FALSE}
library(readr) # this package is for loading files (e.g. tsv, csv, ..)
library(tidyverse) # this package includes ggplot2, dplyr, tidyr, and other tools for data manipulation and visualization
```


# Introduction

The data that we are going to use for this exercise is published in the mBio paper [@niggli2023mbio]. 
It can be found online on: https://doi.org/10.6084/m9.figshare.23578086.v1
It describes the OD600 (growth) of the *Staphylococcus aureus* strain **JE2** in different media.

# Reading in Data and Plotting Growth Curves

```{r load data, eval=TRUE, include=TRUE, echo=TRUE}
dat <- read.table("JE_growth_final.txt", header = TRUE, sep = "\t", stringsAsFactors = TRUE) 
# Load tab-separated data from a text file, treating the first row as column headers and converting character columns to factors
```

```{r print data, eval=TRUE, include=FALSE, echo=TRUE}
print("these are so-called chunks of code, you may include them to be shown in the rendered document (html here). Each chunk has a unique name next to it and then some additional arguments")
print("JE_growth_final.txt")  # Print the filename directly
```

```{r, eval=TRUE, include=TRUE, echo=FALSE}
#Basic Data Inspection
str(dat)       # Structure of the dataset (data types and first few values)
dim(dat)       # Dimensions (number of rows and columns)
head(dat)      # First 6 rows of the dataset
tail(dat)      # Last 6 rows of the dataset
colnames(dat)  # Column names

#Summary Statistics
summary(dat)   # Summary statistics for each column

#Checking for Missing Values
sum(is.na(dat))        # Total number of missing values
colSums(is.na(dat))    # Missing values per column

#Checking Factor Levels (if categorical data exists)
sapply(dat, class)  # Check the data type of each column
lapply(dat, levels) # Show factor levels (if applicable)
```


```{r dimensions, eval=TRUE, include=TRUE, echo=FALSE}
ggplot(dat, aes(x = Time, y = OD600_blank_corrected, group = Growth_Medium, color = Growth_Medium)) +
  geom_line(linewidth = 0.5) +
  geom_point() +
  theme_minimal()
```

```{r}
# Visualize growth of bacteria over time for each Growth Medium
dat %>%
  ggplot(aes(x = Time, y = OD600_blank_corrected, color = Growth_Medium)) +  # Define x-axis as Time, y-axis as OD600, and color by Growth_Medium
  geom_line(stat = "summary", fun = mean, linewidth = 1) +  # Plot the mean OD600 value at each time point (summary statistic)
  geom_point(alpha = 0.2, size = 0.5) +  # Add individual data points with transparency (alpha) for clarity
  labs(
    title = "Bacterial Growth Over Time",  # Set plot title
    x = "Time (hours)",                    # Label for the x-axis
    y = "OD600 (Blank Corrected)",         # Label for the y-axis
    color = "Growth Medium"                # Legend label for the Growth Medium
  ) +
  theme_minimal() +  # Use a simple minimal theme for the plot
  theme(legend.position = "top")  # Position the legend at the top of the plot
```
```{r}
# Visualize growth of bacteria over time for each Growth Medium in separate plots
dat %>%
  ggplot(aes(x = Time, y = OD600_blank_corrected, color = Growth_Medium)) +  # Define x-axis as Time, y-axis as OD600, and color by Growth_Medium
  geom_line(stat = "summary", fun = mean, linewidth = 1) +  # Plot the mean OD600 value at each time point
  geom_point(alpha = 0.2, size = 0.5) +  # Add individual data points with transparency (alpha) for clarity
  labs(
    title = "Bacterial Growth Over Time",
    x = "Time (hours)",
    y = "OD600 (Blank Corrected)",
    color = "Growth Medium"
  ) +
  theme_minimal() +  # Use a simple minimal theme for the plot
  theme(legend.position = "none") +  # Hide the legend (since each plot has a separate color)
  facet_wrap(~ Growth_Medium, scales = "free_y")  # Create separate plots for each Growth Medium with independent y-axis scales
```
```{r}
# Visualize growth of bacteria over time for each Growth Medium and Experiment combination in separate plots
dat %>%
  ggplot(aes(x = Time, y = OD600_blank_corrected, color = Growth_Medium)) +  # Define x-axis as Time, y-axis as OD600, and color by Growth_Medium
  geom_line(stat = "summary", fun = mean, linewidth = 1) +  # Plot the mean OD600 value at each time point
  geom_point(alpha = 0.2, size = 0.5) +  # Add individual data points with transparency (alpha) for clarity
  labs(
    title = "Bacterial Growth Over Time by Experiment and Growth Medium",
    x = "Time (hours)",
    y = "OD600 (Blank Corrected)",
    color = "Growth Medium"
  ) +
  theme_minimal() +  # Use a simple minimal theme for the plot
  theme(legend.position = "none") +  # Hide the legend (since each plot has a separate color)
  facet_grid(Growth_Medium ~ Experiment, scales = "free_y")  # Facet by both Growth_Medium and Experiment, with independent y-axis scales
```

```{r}
  dat %>%
    ggplot(aes(x = Time, y = OD600_blank_corrected, color = Experiment, linetype = Growth_Medium, group = interaction(Strain, Growth_Medium, Replicate))) +
    geom_point(linewidth = 2, alpha = 0.8) +  # Add points for each measurement
    labs(
      title = "Bacterial Growth Over Time",
      x = "Time (Hours)",
      y = "OD600 (Corrected)",
      color = "Strain",
      linetype = "Growth Medium"
    ) +
    theme_minimal() +
        facet_wrap(~ Growth_Medium) +
    theme(legend.position = "right")

```

#Hypothesize (or test) if there is a significant difference for growing the cultures in these three different media
```{r}
# Perform one-way ANOVA to compare OD600 across different Growth Media
anova_results <- dat %>%
  aov(OD600_blank_corrected ~ Growth_Medium, data = .)

# Summarize ANOVA results
summary(anova_results)

# If ANOVA is significant, perform a post-hoc test (Tukey's HSD)
post_hoc_results <- TukeyHSD(anova_results)

# Display post-hoc results
summary(post_hoc_results)
```

#heatmap on quantitative matrix of protein abundances (labels and proteins are blinded)
```{r}
# Use read.table() for tab-separated files
protein_data <- read.table("qmat_1198.txt", sep = "\t", header = TRUE)

# View the first few rows of the dataset
head(protein_data)

# Check the structure of the dataset
str(protein_data)

# Reshape the data to long format (Protein_Gene is the identifier, and the samples are columns)
protein_data_long <- protein_data %>%
  pivot_longer(cols = starts_with("sample"),  # Select columns that start with "sample"
               names_to = "Sample",          # New column for sample names
               values_to = "Abundance")     # New column for protein abundance values

# View the reshaped data
head(protein_data_long)

```

```{r}
library(gplots)  # Loads the 'gplots' package, which contains the heatmap.2 function
```

```{r}
# Convert the data frame to a matrix, excluding the first column (assuming it contains labels)
mat <- as.matrix(protein_data[,2:ncol(protein_data)])  

# Replace all zero values in the matrix with NA (to avoid taking log of zero)
mat[mat == 0] <- NA  

# Apply log2 transformation to the matrix, removing any NA values before transformation
logmat <- log2(na.omit(mat))  

# Generate a heatmap using the transformed matrix, disabling trace lines for clarity
heatmap.2(logmat, trace="none")  

```




# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


# References {-}

