---
title: "Plotting Growth Curves"
author: "UZH n UU: Who is the author here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=TRUE, echo=FALSE}
library(readr) # this package is for loading files (e.g. tsv, csv, ..)

```


# Introduction

The data that we are going to use for this exercise is published in the mBio paper [@niggli2023mbio]. 
It can be found online on: https://doi.org/10.6084/m9.figshare.23578086.v1
It describes the OD600 (growth) of the *Staphylococcus aureus* strain **JE2** in different media.

# Reading in Data and Plotting Growth Curves

```{r load data, eval=TRUE, include=TRUE, echo=TRUE}
fN <- "JE_growth_final.txt" # us this txt file that is copy pasted spread sheet from excel tab separated 
dat <- read.table(fN, header = T, sep = "\t", stringsAsFactors = TRUE)
```

```{r print data, eval=TRUE, include=FALSE, echo=TRUE}
# this print command is only shown on the html if include = TRUE
print("these are so called junks of code, you may include them to be shown in the rendered document (html here) each junk has a unique name next to it and then some additional arguments")
print(fN)
```

```{r dimensions, eval=TRUE, include=TRUE, echo=FALSE}
# 
library(tidyverse)
dd <- dim(dat)
str(dat)
summary(dat)
head(dat)



ggplot(dat, aes(x = Time, y = OD600_blank_corrected, group = Growth_Medium, color = Growth_Medium)) +
  geom_line(linewidth = 0.5) +
  geom_point() +
  theme_minimal()
```

```{r}

  dat %>%
    ggplot(aes(x = Time, y = OD600_blank_corrected, color = Experiment, linetype = Growth_Medium, group = interaction(Strain, Growth_Medium, Replicate))) +
    geom_point(linewidth = 2, alpha = 0.8) +  # Add points for each measurement
    labs(
      title = "Bacterial Growth Over Time",
      x = "Time (Hours)",
      y = "OD600 (Corrected)",
      color = "Strain",
      linetype = "Growth Medium"
    ) +
    theme_minimal() +
        facet_wrap(~ Growth_Medium) +
    theme(legend.position = "right")

```

```{r}
# Call function to generate plot
plot_growth_curve(dat)
```



We will:
- Load and explore the OD600 dataset
- Plot growth curves using `ggplot2`
- Create a heatmap for protein abundances



```{r dimensions, eval=TRUE, include=TRUE, echo=FALSE}
## **1. Load Required Libraries**


# Load necessary packages
library(ggplot2)    # For plotting
library(dplyr)      # For data manipulation
library(tidyr)      # For reshaping data
library(pheatmap)   # For heatmap visualization

# Create heatmap of protein abundances
pheatmap(protein_data,
         scale = "row",         # Normalize rows
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         color = colorRampPalette(c("blue", "white", "red"))(100), main = "Protein Abundance Heatmap")
```
#

The dimensions of this dataset are shown above.
It is easy possible to show also content of variables in the html part.  
Note: if you end a line with 2 spaces you go to a new paragraph.  
  
Like here: `r dd`


# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


# References {-}

